Topological Equivariant Artist Model for Visualization Library Architecture

### 2023-02-08 paper/main.tex
finished out RW top section

\section{Related Work}
We build on the extensive work codifying the structure of visualization data to develop a more generalizable method for expressing the structure that visualizations are expected to preserve. Structure preservation is important because it means visualzations are easier to understand \cite{norman_things_smart} and ensures that the visual representations are not distortions of the data\cite{tufteVisualDisplayQuantitative2001}. Broadly, previous work describes structure as a combination of the topological properties of the data, the mathematical structure of the data fields, and equivalent or compatible changes to data and graphics. We propose that a unified abstraction model can guide developers in building structure aware composable software library components with more consistent interfaces.

@@

A topology is a partitioning of a space into sets of its subspaces and which connectivity is encoded in the topology is determined by the way in which the space is partitioned. For example, the GHCN\cite{lawrimoreGlobalHistoricalClimatology2011} dataset in \autoref{fig:related-work:continuity:ktypes} is a complex spatio-temporal record of weather station data. One way of encoding that each record is independent is to encode the underlying topology as a set of discrete points, one for each record. Each of these discrete points has a corresponding marker in the scatter plot. Another way of subsetting the data is as sets of timeseries, which can be encoded as intervals since weather station data is sampled from 1D continous natural phenomena. Each line plot can be thought of as a function evaluated on the underlying interval that encoding the temporal continuity. A third way to view the data is as a sparse sampling of a 2D geospatial grid, where the connectivity can be modeled as a 2D plane. Here the connectivity is preserved by plotting the stations in the map such that their relative distances are preserved.

note{add summary and transition}

Visual algorithms assume the topology of their input data, as described in taxonomies of visualization algorithms Chi\cite{chiTaxonomyVisualizationTechniques2000} and by Troy and M\"{o}ller \cite{toryRethinkingVisualizationHighlevel2004}. For example, a \texttt{line} algorithm often does not have a way to query whether a list of (x,y) coordinates is the distinct rows, the time series, or the list of stations in \autoref{fig:related-work:continuity:ktypes}. While plotting the time series as a continuous line would be correct, it would be incorrect for a visualization to indicate that the distinct rows or stations are connected in a 1D continuous manner. Since topological continuity is generalizable, we propose the construction of topology types such that data can explicitly state its topology type and visual algorithms can check that this type matches their assumptions. The typing system proposed in this paper builds on Butler's proposal of using a mathematical structure called fiber bundles as an abstract data representation in visualization \cite{butlerVectorBundleClassesForm1992, butlerVisualizationModelBased1989}. We extend Butler's work because he discusses how bundles can be used to encode data in a consistent manner for the topological and field structures that are common in visualization. We sketch out fiber bundles in \autoref{sec:atct:fiber-bundles}, but his work provides a thorough introduction to bundles for visualization practitioners.

>>>>

\section{Related Work}
We build on the extensive work codifying the structure of visualization data to develop a more generalizable method for expressing the structure that tools are expected to preserve when transforming data into visualizations. Structure preservation ensures that visual representations are not distortions of the data\cite{tufteVisualDisplayQuantitative2001} and it makes visualizations easier to understand \cite{norman_things_smart}. Broadly, previous work describes structure as a combination of the topological properties of the data, the mathematical structure of the data fields, and equivalent or compatible changes to data and graphics. We propose that a unified abstraction model can guide developers in building structure aware composable software library components with more consistent interfaces.

@@

A topology is a partitioning of a space into its subspaces in a way where those subspaces can be composed back into the space and the choice of partitions is how we encode the connectivity of the data that we are interested in. For example, the GHCN\cite{lawrimoreGlobalHistoricalClimatology2011} dataset in \autoref{fig:related-work:continuity:ktypes} is a complex spatio-temporal record of weather station data. One way of encoding that each record is independent is to encode the underlying topology as a set of discrete points, one for each record. Each of these discrete points is also represented visually as a discrete marker in the scatter plot. Another way of viewing the data is as sets of timeseries since weather station data is sampled from 1D continous natural phenomena. We encode this topological structure as intervals and sets of sub-intervals, and  we visually represent this strucure as continuous line plots. A third way of viewing the data is as a sparse sampling of a 2D geospatial grid, where the connectivity is modeled as a  surface and sub regions of that surface. In the visualization, that connectivity is expressed as plotting the stations in the map, visually denoting that these are points in the continuous geographical space that is New York State. By using topology as the underlying encoding scheme, we are able to represent the structure of points, lines, surfaces, and more using the same abstractions.

Since topological continuity is generalizable, we propose that encoding topology as part of a dataset type and a visualization type allows us to explicitly check which types of structure a visualization algorithm can preserve. Visual algorithms assume the topology of their input data, as described in taxonomies of visualization algorithms Chi\cite{chiTaxonomyVisualizationTechniques2000} and by Troy and M\"{o}ller \cite{toryRethinkingVisualizationHighlevel2004}, but generally do not verify that input structure. For example, a \texttt{line} algorithm often does not have a way to query whether a list of (x,y) coordinates is the distinct rows, the time series, or the list of stations in \autoref{fig:related-work:continuity:ktypes}. While plotting the time series as a continuous line would be correct, it would be incorrect for a visualization to indicate that the distinct rows or stations are connected in a 1D continuous manner. Encoding the expected input structure would enable a visualization algorithm to reject data that does not explicitely state that it matches those assumptions. This structure can be encoded in a generalizable way by building on Butler's work on using a mathematical structure called fiber bundles as an abstract data representation in visualization \cite{butlerVectorBundleClassesForm1992, butlerVisualizationModelBased1989} where he discusses how bundles can be used to encode data in a consistent manner for the topological and field structures that are common in visualization. We sketch out fiber bundles in \autoref{sec:atct:fiber-bundles}, but his work provides a thorough introduction to bundles for visualization practitioners.
